The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2020-08-31 Writeup/manuscript.tex
Results almost done, Discussion halfway done

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \cite{RN1425}. Wallpaper groups differ in the number and kind of these transformations. In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \cite{RN1425}. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the outer group \cite{RN1425}. As an example, let us consider groups P6 and P2. If we ignore the translations in two directions that both groups share, group P6 consists of the set of rotations {$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$}, in which P2 {$0^{\circ}$, $180^{\circ}$} is contained. P2 is thus a subgroup of P6, and the full P6 set can be generated by every combination of P2 and rotations {$0^{\circ}$, $120^{\circ}$, $240^{\circ}$}. Because P2 is repeated three times in P6, P2 is a subgroup of P6 with index 3 \cite{RN1425}. The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are sub-groups of more complex ones \cite{RN1711}.

\begin{figure}[tb]
\centering
\includegraphics[width=1\linewidth]{../figures/figure1.png}
\caption{Examplar stimulus for each of the 16 wallpaper groups, with corresponding <something EEG> and a boxplot showing the distribution of display duration thresholds. }
\label{fig:durations_rotations}
\end{figure}

@@

The psychophysical experiment took a distinct but related approach. In each trial an exemplar image was shown with its matched control, one image after the other, and the order varied pseudo-randomly such that in half the trials the original exemplar was shown first, and in the other half the control image was shown first. After each trial, participants were told to indicate whether the first or second image contained more structure, and the duration of both images was controlled by a staircase procedure so that a threshold duration could be computed for each wallpaper group.

A summary of our results can be seen in Figure 1 ...

\begin{figure}[tbhp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean RMS EEG response. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that xx/64 subgroup relationships have $p(\Delta|data)>0.9$.}
\label{fig:eeg_comp}
\end{figure}

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \cite{RN1170} and rotation \cite{RN1725}, the visual system explicitly represents hierarchical structure of the 17 wallpaper groups, and thus the compositions of all four of the fundamental symmetry transformations (rotation, reflection, translation, glide reflection) which comprise regular textures. The RMS measure of SSVEP amplitude, preserves the complex hierarchy of subgroup relationships among the wallpaper groups \cite{RN1711}. Out of a total of 60 relationships, 53 were preserved in a significant number of participants, and 49 were significant even at a stricter threshold (p < 0.002). The ordering was highly stable in individual participants, with an average preservation rate of ~21 of 25 participants across all 60 relationships (see Figure 3). This remarkable consistency was specific to the odd harmonics of the stimulus frequency, that capture the symmetry-specific response \cite{RN1725} and to electrodes in an ROI over occipital cortex. When the same analysis was done on the even harmonics of the occipital cortex ROI, the ordering of responses was much less apparent (see Figure S2) and preservation rates much lower (see Figure S4). The odd harmonics from electrodes in an ROI over parietal cortex, showed even weaker evidence of preserving the hierarchy among sub-groups (see Figure S5). Importantly, no relationships were preserved in either of these control analyses that were not also preserved in the main analysis of the odd harmonics in the occipital cortex ROI. The current data provide a complete description of the visual system’s response to symmetries in the 2-D plane. Our design does not allow us to independently measure the response to P1, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 2), we can conclude that the relationships between P1 and all other groups, where P1 is the subgroup, are also preserved by the visual system. The subgroup relationships are not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. Furthermore, behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \cite{RN1253}, they are generally error-prone when asked to assign exemplar images to the appropriate group \cite{RN172}. The correspondence between responses in the visual system and group theory that we demonstrate here, may reflect a form of implicit learning that depends on the structure of the natural world. The environment is itself constrained by physical forces underlying pattern formation and these forces are subject to multiple symmetry constraints \cite{RN1634}. The ordered structure of responses to wallpaper groups could be driven by a central tenet of neural coding, that of efficiency. If coding is to be efficient, neural resources should be distributed in such a way that the structure of the environment is captured with minimum redundancy considering the visual geometric optics, the capabilities of the subsequent neural coding stages and the behavioral goals of the organism \cite{RN1758, RN1760, RN1757, RN1756}. Early work within the efficient coding framework suggested that natural images had a $1/f$ spectrum and that the corresponding redundancy between pixels in natural images could be coded efficiently with a sparse set of oriented filter responses, such as those present in the early visual pathway \cite{RN1740, RN1446}. Our results suggest that the principle of efficient coding extends to a much higher level of structural redundancy – that of symmetries in visual images. The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \cite{RN1519}. The correspondence between brain data and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \cite{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \cite{RN1768}. Field biologist have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \cite{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered super-textures, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \cite{RN1764}. Wallpaper groups are visually compelling and have been widely used in human artistic expression going back to the Neolithic age \cite{RN1949}. If wallpapers are super-textures, their prevalence may be a direct consequence of the strategy the human visual system uses for encoding visual textures.

\begin{figure}%[tbhp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{}}
\label{fig:durations_rotations}
\end{figure}

@@

\subsection*{EEG Analysis}
Response waveforms were generated for each group by selective filtering in the frequency domain. For each participant, the average Fourier coefficients from the two sessions were averaged over trials and sessions. The Steady-State Visual Evoked Potentials paradigm we used allowed us to separate symmetry-related responses from non-specific contrast transient responses. Previous work has demonstrated that symmetry-related responses are predominantly found in the odd harmonics of the stimulus frequency, whereas the even harmonics consist mainly of responses unrelated to symmetry, that arise from the contrast change associated with the appearance of the second image[2-4]. This functional distinction of the harmonics allowed us to generate a single-cycle waveform containing the response specific to symmetry, by filtering out the even harmonics in the spectral domain, and then back-transforming the remaining signal, consisting only of odd harmonics, into the time-domain. For our main analysis, we averaged the odd harmonic single-cycle waveforms within a six-electrode region of interest (ROI) over occipital cortex (electrodes 70, 74, 75, 81, 82, 83). These waveforms, averaged over participants, are shown in Figure 2 in the main paper. The same analysis was done for the even harmonics (see Figure S1) and for the odd harmonics within a six electrode ROI over parietal cortex (electrodes 53, 54, 61, 78, 79, 86; see Figure S2). The root-mean square values of these waveforms, for each individual participant, were used to determine whether each of the wallpaper subgroup relationships were preserved in the brain data.

>>>>

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \cite{RN1425}. Wallpaper groups differ in the number and kind of these transformations. In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \cite{RN1425}. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the outer group \cite{RN1425}. As an example, let us consider groups P6 and P2. If we ignore the translations in two directions that both groups share, group P6 consists of the set of rotations {$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$}, in which P2 {$0^{\circ}$, $180^{\circ}$} is contained. P2 is thus a subgroup of P6, and the full P6 set can be generated by every combination of P2 and rotations {$0^{\circ}$, $120^{\circ}$, $240^{\circ}$}. Because P2 is repeated three times in P6, P2 is a subgroup of P6 with index 3 \cite{RN1425}. The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are sub-groups of more complex ones \cite{RN1711}. The full set of subgroup relationships is listed in Section 1.4.2 of the Supplementary Material.

\begin{figure}[tb]
\centering
\includegraphics[width=1\linewidth]{../figures/figure1.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a P1 exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:durations_rotations}
\end{figure}

@@

The psychophysical experiment took a distinct but related approach. In each trial an exemplar image was shown with its matched control, one image after the other, and the order varied pseudo-randomly such that in half the trials the original exemplar was shown first, and in the other half the control image was shown first. After each trial, participants were told to indicate whether the first or second image contained more structure, and the duration of both images was controlled by a staircase procedure so that a display duration threshold could be computed for each wallpaper group.

A summary of our brain imaging and psychophysical measurements is presented with examples of the wallpaper groups in Figure 1. For our primary SSVEP analysis, we only considered EEG data from a pre-determined region-of-interest (ROI) consisting of six electrodes over occipital cortex (see Supplementary Figure 1.1). SSVEP data from this ROI was filtered so that only the odd harmonics that capture the symmetry response contribute to the waveforms. While waveform amplitude is quite variable among the 16 groups, all groups have a sustained negative-going response that begins at about the same time for all groups, ~180 ms after the transition from the P1 control exemplar to the original exemplar. To reduced the amplitude of the symmetry-specific response to a single number that could be used in further analyses and compared to the psychophysical data, we computed the root-mean-square (RMS) over the odd-harmonic-filtered waveforms. The data in Figure 1 are shown in descending order according to RMS. The psychophysical results, shown in box plots in Figure 1, were also quite variable between groups, and there seems to be a general pattern where wallpaper groups near the top of the figure, that have lower SSVEP amplitudes, also have longer psychophysical display duration thresholds.

We now wanted to quantify the degree to which our two measurements were consistent with the subgroup relationships. We hypothesized that more complex groups would (a) produce symmetric-specific SSVEPs with higher amplitudes and (b) shorter display duration thresholds. [PERHAPS WE NEED TO JUSTIFY THESE ASSUMPTIONS] We tested each of these hypotheses using the same approach. We first fitted a Bayesian model with wallpaper group as a factor and participant as a random effect. We fit the model separately for SSVEP RMS and psychophysical data, and then computed posterior distributions for the difference between supergroup and subgroup. These difference distributions could allowed us to compute the conditional probability that the supergroup would produce (a) larger RMS and (b) a shorter threshold durations, when compared to the subgroup. The posterior distributions are shown in Figure 2 for the SSVEP data, and in Figure 3 for the psychophysical data, which distributions color-coded according to conditional probability. For both data sets our hypothesis is confirmed: For the overwhelming majority of the 64 subgroup relationships, supergroups are more likely to produce larger symmetry specific SSVEPs and shorter threshold durations, and in most cases the conditional probability of this happening is extremely high.

We also ran a control analysis using (1) odd-harmonic SSVEP data from a six-electrode ROI over parietal cortex (see Supplementary Figure 1.1) and (2) even-harmonic SSVEP data from the same occipital ROI that was used in our primary analysis. By comparing these two control analysis to our primary SSVEP analysis, we can address the specify of our effects in terms of location (occipital cortex vs parietal cortex) and harmonic (odd vs even). For both control analyses (plotted in Supplementary Figures 3.3 and 3.4), the correspondence between data and subgroup relationships was weaker than in the primary analysis. We can quantify the strength of the assocation between the data and the subgroup relationships, by asking what proportion of subgroup relationships that reach or exceed a range of probability thresholds. This is plotted in Figure 4, for our psychophysical data, our primary SSVEP analysis and our two control SSVEP analyses. It it that odd-harmonic SSVEP data from an occipital ROI and display duration thresholds both have a strong association with the subgroup relations, that for a clear majority of the subgroups survive even at the highest threshold we consider ($p(\Delta|data)>0.99$), and that the assocation is far weaker for the two control analyses.

A subset of the SSVEP data was previously published as part of an earlier demonstration of parametric responses to rotation symmetry in wallpaper groups \cite{RN1725}. We replicate that result using our Bayesian approach, and find the same parametric effect in the psychophysical data (Supplementary Figure 4.1 - NEEDS TO BE LABELED IN THE SM). We also conducted analyses looking for effects of index and normality in our two datasets.

We conclude that the hierarchy of complexity among wallpaper groups is reflected in precise detail in both brain responses and performance in a psychophysical task.

\begin{figure}[tbhp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean RMS SSVEP response. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that xx/64 subgroup relationships have $p(\Delta|data)>0.9$.}
\label{fig:eeg_comp}
\end{figure}

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \cite{RN1170} and rotation \cite{RN1725}, the visual system explicitly represents hierarchical structure of the 17 wallpaper groups, and thus the compositions of all four of the fundamental symmetry transformations (rotation, reflection, translation, glide reflection) which comprise regular textures. The RMS measure of SSVEP amplitude, preserves the complex hierarchy of subgroup relationships among the wallpaper groups \cite{RN1711}. This remarkable consistency was specific to the odd harmonics of the stimulus frequency, that capture the symmetry-specific response \cite{RN1725} and to electrodes in an ROI over occipital cortex. When the same analysis was done using the odd harmonics from a parietal cortex ROI (Supplementary Figure 3.4) or using the even harmonics of the occipital cortex ROI (Supplementary Figure 3.4), the data was much less consistent with the subgroup relationships (yellow and green lines, Figure 4).

The current data provide a complete description of the visual system’s response to symmetries in the 2-D plane. Our design does not allow us to independently measure the response to P1, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 1), we can conclude that the relationships between P1 and all other groups, where P1 is the subgroup, are also preserved by the visual system. The subgroup relationships are not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. Furthermore, behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \cite{RN1253}, they are generally error-prone when asked to assign exemplar images to the appropriate group \cite{RN172}. The correspondence between responses in the visual system and group theory that we demonstrate here, may reflect a form of implicit learning that depends on the structure of the natural world. The environment is itself constrained by physical forces underlying pattern formation and these forces are subject to multiple symmetry constraints \cite{RN1634}. The ordered structure of responses to wallpaper groups could be driven by a central tenet of neural coding, that of efficiency. If coding is to be efficient, neural resources should be distributed in such a way that the structure of the environment is captured with minimum redundancy considering the visual geometric optics, the capabilities of the subsequent neural coding stages and the behavioral goals of the organism \cite{RN1758, RN1760, RN1757, RN1756}. Early work within the efficient coding framework suggested that natural images had a $1/f$ spectrum and that the corresponding redundancy between pixels in natural images could be coded efficiently with a sparse set of oriented filter responses, such as those present in the early visual pathway \cite{RN1740, RN1446}. Our results suggest that the principle of efficient coding extends to a much higher level of structural redundancy – that of symmetries in visual images. The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \cite{RN1519}. The correspondence between brain data and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \cite{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \cite{RN1768}. Field biologist have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \cite{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered super-textures, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \cite{RN1764}. Wallpaper groups are visually compelling and have been widely used in human artistic expression going back to the Neolithic age \cite{RN1949}. If wallpapers are super-textures, their prevalence may be a direct consequence of the strategy the human visual system uses for encoding visual textures.

\begin{figure}[tbhp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{}}
\label{fig:durations_rotations}
\end{figure}

@@

\subsection*{SSVEP Analysis}
Response waveforms were generated for each group by selective filtering in the frequency domain. For each participant, the average Fourier coefficients from the two sessions were averaged over trials and sessions. The Steady-State Visual Evoked Potentials (SSVEP) paradigm we used allowed us to separate symmetry-related responses from non-specific contrast transient responses. Previous work has demonstrated that symmetry-related responses are predominantly found in the odd harmonics of the stimulus frequency, whereas the even harmonics consist mainly of responses unrelated to symmetry, that arise from the contrast change associated with the appearance of the second image[2-4]. This functional distinction of the harmonics allowed us to generate a single-cycle waveform containing the response specific to symmetry, by filtering out the even harmonics in the spectral domain, and then back-transforming the remaining signal, consisting only of odd harmonics, into the time-domain. For our main analysis, we averaged the odd harmonic single-cycle waveforms within a six-electrode region of interest (ROI) over occipital cortex (electrodes 70, 74, 75, 81, 82, 83). These waveforms, averaged over participants, are shown in Figure 2 in the main paper. The same analysis was done for the even harmonics (see Figure S1) and for the odd harmonics within a six electrode ROI over parietal cortex (electrodes 53, 54, 61, 78, 79, 86; see Figure S2). The root-mean square values of these waveforms, for each individual participant, were used to determine whether each of the wallpaper subgroup relationships were preserved in the brain data.
